Open-source tools in R for landscape ecology

### 2021-04-24 Comments_Reviewer.Rmd
:construction_worker: Starting to address the first reviewer comments

@@

@@

"Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, R allows for transparent and reproducible workflows."

But the authors do not mention any tools for facililitating reproducibility.
This is a huge part of why R should be more widely used, and should be expanded upon.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

**Reply**: We agree with the reviewer that this sentence might be misleading. We tried to reword the statement in a way that makes clear base R does not have these capabilities, but many package provide it. \vskip 0.5cm

@@

**Reply**: We added that we aware that the survey is most likely not representative for the R community and R users due to how we tried to reach the community and also due to the questions asked in the survey. Since the survey is not the main idea of the manuscript though, we think that the survery still migth be interesting even with these shortcomings. \vskip 0.5cm

@@

**Reply**: We agree that the sentence implied that the R community is not constantly adapting to the new developments of landscape ecology which is certainly not true. We reworded the sentence making clearea that indeed the community constantly improves the R programming language. However, we do not agree with the reviewer that "[...] the problem is that there are so many packages [..]". Even though we acknowledge that the amount of package can be overwhelming, especially for novices users, diversity of packages is a big strength - even for similar specific tasks. Possible advantages include e.g., the possibility to compare different implementations of similar/identical task or providing functionality for slightly different data structures. \vskip 0.5cm

"Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, R allows for transparent and reproducible workflows."
But the authors do not mention any tools for facililitating reproducibility.
This is a huge part of why R should be more widely used, and should be expanded upon.

**Reply**: We highly agreed that this one of the main reasons why R should be widely used and we added an explanation how the script-based character of R itself allows to share and reproduce analysis. \vskip 0.5cm

@@

**Reply**: Please see our reply to the previous comment. We reworded the sentence to state clearer that this is indeed one of the strength of the R community. \vskip 0.5cm

@@

**Reply**: We added some explanation that raster and terra already provide solutions to this issue by processing only chunks that can be stored in memory. Additionally, we added some information about parallel processing as another possible solution to these issues. However, we believe that discussing underlying issues of R memory management is outside the scope if this paper. \vskip 0.5cm

@@

**Reply**: We apologise for this. The manuscript was written using the rticles packages with the corresponding template for Springer journals which somehow did not include journal names. We fixed the citation style including journal names. \textcolor{red}{STILL NEED TO DO THIS ACTUALLY!} \vskip 0.5cm

@@

**Reply**: We agree with the reviewer that the provided script is not used for experienced R users and targeted to novices.
Thus, we tried to improve the the documentation of the code. \vskip 0.5cm

@@

**Reply**: We agree that the use of viridis might be confusing without mentioning it in the manuscript and we removed it from the R script.
We agree that ggplot2 graphics are a very large field.
This also highlights that it is outside the scope of this paper.
Nevertheless, ggplot2 is used quite frequently by the R community, so we decided to include it in the example.
We added a reference to the manuscript pointing towards ggplot2 specific resources. \vskip 0.5cm

@@

**Reply**: We agree with the reviewer and hope that by trying to address the very helpful and valuable comments by the reviewers, we approach this goal to a higher degree. \vskip 0.5cm

@@

**Reply**: Thank you very much for the kind and encouraging words about the manuscript.
We are happy to hear that the reviewer thinks the manuscript might be useful for ecologists that would like to analyze data through R. \vskip 0.5cm

@@

**Reply**: Since both the section editor and reviewer 1 suggested to remove the definition of landscape ecology due to the target audience of the journal, the introduction was rewritten and shorted including this phrase. \vskip 0.5cm

@@

**Reply**: We changed the sentence as suggested. \vskip 0.5cm

@@

**Reply**: We agree that many users are used to devtools to install packages from GitHub.
However, we decided to only mention remotes in this context since it is the more recent replacement for this task as also stated on the remotes homepage ("[...] This package is a lightweight replacement of the **install\_** functions in devtools. Indeed most of the code was copied over from devtools [...]" <https://github.com/r-lib/remotes#readme>) \vskip 0.5cm

@@

**Reply**: We strongly agree with the reviewer that this information would be very valuable to judge the representation of the survey.
However, unfortunately such questions were not included.
We also acknowledge that the survey is most likely not representative for all skill levels of *R* users because we assume that by using mailing list and social media more advanced users were more likely to take part in the survey.
We added and clearified those shortcomings of the survey. \vskip 0.5cm

@@

**Reply**: The GitHub repository is currently set to "private" and will be publicly available upon publication. \vskip 0.5cm

### 2021-04-24 paper/paper.Rmd
:construction_worker: Starting to address the first reviewer comments

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}.
Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g. Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The *R* programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig. \@ref(fig:fig-workflow)).
The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.

@@

Most *R* packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
There are three major online platforms to host *R* packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
The last one focuses on tools for the analysis of genomic data; therefore, we focus on only the former two in this review.
CRAN (the *Comprehensive R Archive Network*) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
*GitHub* hosts source code under version control, and allows users to install packages with one line of code using the *remotes* \cite{Hester2020} package.
Additionally, hosting a package on *GitHub* provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e. testing if functions return an expected value).

@@

While *R* has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL/OGRcontributors2020} and *PROJ* \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
Both, *tmap* (Fig. \@ref(fig:fig-map)c) and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

@@

Most landscape metrics are a represented by a single number depicting specific characteristics of a local landscape.
Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram \cite{Nowosad2021}.
Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
All of the spatial signatures methods mentioned above are implemented in the *motif* package \cite{Nowosad2021}.

@@

Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across various *R* packages.
A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the *NLMR* package \cite{Sciaini2018}.
Furthermore, the *RandomFields* package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.

@@

To better understand how the landscape ecology community uses *R*, we conducted a short survey using mailing lists and social media to reach the community. 
All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).

In total, the survey was answered by `r n_people` participants, of which the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
Other, less frequent answers were "Data scientists", "None of the above", "Government employees" "Master's degree student" and "Bachelor's degree student" (in decreasing order).

@@

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey)b).
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").

@@

Since its first introduction in 1995, *R* has come a long way from an exclusively statistical programming language to a powerful landscape ecology tool.
Today, many *R* packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed at spatial data handling and analysis.
The highly dynamic development of *R* packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
However, since landscape ecology constantly develops and improves, the *R* programming language and its packages need to change and adapt to these changes.

A comprehensive collection of *R* software packages exists to handle the most common tasks of landscape ecology. 
Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, *R* allows for transparent and reproducible workflows.
Additionally, this also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

@@

Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
While the *R* programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the *R* programming language has to face. 
Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the *R* packages related to those fields.
But as the development of *R* packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.

>>>>

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}. 
\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g. Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The *R* programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig. \@ref(fig:fig-workflow)).
The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.

@@

Most *R* packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
There are three major online platforms to host *R* packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two".}

CRAN (the *Comprehensive R Archive Network*) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
*GitHub* hosts source code under version control, and allows users to install packages with one line of code using the *remotes* \cite{Hester2020} package.
Additionally, hosting a package on *GitHub* provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e. testing if functions return an expected value).

@@

While \textcolor{blue}{base *R*} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data. 
\textcolor{blue}{However, as discussed previously, one strength and core idea of the *R* programming language are its expandability by pacakges.} 
Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL/OGRcontributors2020} and *PROJ* \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}. 
\textcolor{blue}{Fore more information about **ggplot2** in general see \cite{Wickham2016}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
Both, *tmap* (Fig. \@ref(fig:fig-map)c) and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

@@

Most landscape metrics are represented by a single number depicting specific characteristics of a local landscape.
Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram \cite{Nowosad2021}.
Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
All of the spatial signatures methods mentioned above are implemented in the *motif* package \cite{Nowosad2021}.

@@

Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across \textcolor{blue}{a couple} *R* packages.
A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the *NLMR* package \cite{Sciaini2018}.
Furthermore, the *RandomFields* package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.

@@

To better understand how the landscape ecology community uses *R*, we conducted a short survey using mailing lists and social media to reach the community. \textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of *R* users and results are most likely biased towards more advanced users. Also, the survey did not include any personal questions, such as age or country of residence, which might have an effect on the representation of the survey.}.
All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).

In total, the survey was answered by `r n_people` participants, of which the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
Other, less frequent answers were \textcolor{blue}{"Data scientists", "None of the above", "Government employees", "Master's degree student" and "Bachelor's degree student" (in decreasing order).}

@@

Next, we were interested in the most important tasks to the workflow of the participants.
\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey)b).}
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").

@@

Since its first introduction in 1995, *R* has come a long way from an exclusively statistical programming language to a powerful landscape ecology tool.
Today, many *R* packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed at spatial data handling and analysis.
The highly dynamic development of *R* packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
However, since landscape ecology constantly develops and improves, \textcolor{blue}{consequently also the *R* programming language and its packages are under constant change to adapt to these new developments.}

A comprehensive collection of *R* software packages exists to handle the most common tasks of landscape ecology.
Because it is possible to import, modify, analyze, and visualize spatial data all in the same  \textcolor{blue}{scripted} programming environment, *R* allows for transparent and reproducible workflows.
\textcolor{blue}{This script-based characteristic of the *R* programming language allows to easily share and reproduce analysis, as demonstrated by the R script in the appendix, which is a big advantage over point-and-click software interfaces.}
This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

@@

Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
While the *R* programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the *R* programming language has to face. 
\textcolor{blue}{Some packages discussed herein, such as *raster* and *terra*, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
Furthermore, also parallel processing, and even including the use of high performance clusters, is provided by several *R* packages. 
While this is outside the scope of this manuscript, further resources about parallel processing in *R* can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.} 
Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the *R* packages related to those fields.
But as the development of *R* packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
